---
title: 'Machine Learning - Capstone project'
author: "Ashwini Bhatte"
date: "22/08/2017"
---


The dataset is available [here](https://archive.ics.uci.edu/ml/datasets/Diabetes)

**Question**

Can predict blood glucose concentration and  Hyperglycemia (high blood glucose) or Hypoglycemia (low blood glucose) associated with it in diabetes patients.

Following packages are used for our analysis

```{r, message=FALSE, warning=FALSE, include=FALSE}
#install.packages("caTools")
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(caTools)
library(ROCR)
library(nnet)
```
```
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(caTools)
library(ROCR)
library(nnet)
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
# create an empty data frame
df <- data.frame()
#for loop for data files 1:70
for(i in 1:70) {
  num = paste("", i, sep = "")
# if i < 10, paste 0 followed by the file name, e.g 01, 02, etc.
  if (i < 10) num = paste("0", i, sep = "")
  fname = paste("~/R Projects/diabetes-dataset/Diabetes-Data/data-", num, sep = "")
# creat vector to read files and add column names  
  temp <- read_delim(fname, col_names = c("date", "time", "code", "bg_conc"), 
   col_types = cols(
    date = col_date(format = "%m-%d-%Y"),
# %R = Equivalent to %H:%M.    
    time = col_time(format = "%R"),
# original string converted to integer
    code = col_integer(),
    bg_conc = col_character()
  ), "\t",
    escape_double = FALSE, trim_ws = TRUE);
# add patient_num column to represent each patients records (1:70)
  temp <- mutate(temp, patient_num = i)
# bind data files (1:70) & create df file
  df <- rbind(df, temp);
}

## create clean_df to clean the data

# remove "000" & "3A" from bg_conc as they are illogical values
clean_df <- subset(df, bg_conc != "000" & bg_conc != "3A")
# clean_df contains code "4", "36, & "56", but since there is no entry of such codes in
# the data description, these values are removed as well
clean_df <- subset(clean_df, code != "4" & code != "36" & code != "56")

# For further EDA, convert clean_df$bg_conc from character to numeric

clean_df$bg_conc <- as.numeric(clean_df$bg_conc)
#remove NAs from above
clean_df <- na.omit(clean_df)

# Since time is a continuous variable, it will be difficult to explore the data when time is plotted on x- axis. Which is why the time intervals of 2 hours (time_grp) were created

clean_df <- clean_df %>%
  mutate(time_grp = gsub("^00:[0-9][0-9]:00|^01:[0-9][0-9]:00", "00-02", x = time))%>%
  mutate(time_grp = gsub("^02:[0-9][0-9]:00|^03:[0-9][0-9]:00", "02-04", x = time_grp))%>%
  mutate(time_grp = gsub("^04:[0-9][0-9]:00|^05:[0-9][0-9]:00", "04-06", x = time_grp))%>%   
  mutate(time_grp = gsub("^06:[0-9][0-9]:00|^07:[0-9][0-9]:00", "06-08", x = time_grp))%>%
  mutate(time_grp = gsub("^08:[0-9][0-9]:00|^09:[0-9][0-9]:00", "08-10", x = time_grp))%>%
  mutate(time_grp = gsub("^10:[0-9][0-9]:00|^11:[0-9][0-9]:00", "10-12", x = time_grp))%>%
  mutate(time_grp = gsub("^12:[0-9][0-9]:00|^13:[0-9][0-9]:00", "12-14", x = time_grp))%>%
  mutate(time_grp = gsub("^14:[0-9][0-9]:00|^15:[0-9][0-9]:00", "14-16", x = time_grp))%>%    
  mutate(time_grp = gsub("^16:[0-9][0-9]:00|^17:[0-9][0-9]:00", "16-18", x = time_grp))%>%
  mutate(time_grp = gsub("^18:[0-9][0-9]:00|^19:[0-9][0-9]:00", "18-20", x = time_grp))%>%     
  mutate(time_grp = gsub("^20:[0-9][0-9]:00|^21:[0-9][0-9]:00", "20-22", x = time_grp))%>%
  mutate(time_grp = gsub("^22:[0-9][0-9]:00|^23:[0-9][0-9]:00", "22-24", x = time_grp))

#From BG concentration measurements, we can derive if the patient is Hypoglycemic, Hyperglycemic or has a blood glucose in average range. 

#To study the effects, 3 groups are createrd

#bg_level: 0-80mg/dl = Hypoglycaemic, 81-199mg/dl = Average, above 200mh/dl = Hyperglycemic

clean_df <- clean_df %>%
  mutate(bg_symp = gsub("^([0-9]|[0-7][0-9]|80)$", "Hypoglycemia", x = bg_conc))%>%
  mutate(bg_symp = gsub("^(8[1-9]|9[0-9]|1[0-9][0-9])$", "Average", x = bg_symp))%>%
  mutate(bg_symp = gsub("^(2[0-9][0-9]|3[0-9][0-9]|4[0-9][0-9]|50[0-1])$", 
                        "Hyperglycemia", x = bg_symp)) %>%
  mutate(bg_symp = gsub("^(1.5|2.5|3.5|4.5|6.5|7.5)$", "Hypoglycemia", x = bg_symp))

# Let's add a new column "activity" which represent a type of activity patient has done just
# before or after measuring the BG
# code        type of activity
# 33 - 35       insulin dose
# 58 - 64       meal (pre - post)
# 69 - 71       exercise

clean_df <- clean_df %>%
  mutate(activity = gsub("^(3[3-5])$", "insulin", x = code)) %>%
  mutate(activity = gsub("^(5[8-9]|6[0-4]|66|67|68)$", "meal", x = activity)) %>%
  mutate(activity = gsub("^(6[9]|7[0-1])$", "exercise", x = activity)) %>%
  mutate(activity = gsub("^(48|57|72)$", "unspecified", x = activity)) %>%
  mutate(activity = gsub("^(65)$", "hypo-measurement", x = activity))

# Add one more column similar to the time_bin. 
# This represents time groups in a numeric format e.g. 00-02 time_bin = 0 bin_num, etc.
clean_df <- clean_df %>%
  mutate(bin_num = gsub("^00:[0-9][0-9]:00|^01:[0-9][0-9]:00", "0", x = time))%>%
  mutate(bin_num = gsub("^02:[0-9][0-9]:00|^03:[0-9][0-9]:00", "2", x = bin_num))%>%
  mutate(bin_num = gsub("^04:[0-9][0-9]:00|^05:[0-9][0-9]:00", "4", x = bin_num))%>%   
  mutate(bin_num = gsub("^06:[0-9][0-9]:00|^07:[0-9][0-9]:00", "6", x = bin_num))%>%
  mutate(bin_num = gsub("^08:[0-9][0-9]:00|^09:[0-9][0-9]:00", "8", x = bin_num))%>%
  mutate(bin_num = gsub("^10:[0-9][0-9]:00|^11:[0-9][0-9]:00", "10", x = bin_num))%>%
  mutate(bin_num = gsub("^12:[0-9][0-9]:00|^13:[0-9][0-9]:00", "12", x = bin_num))%>%
  mutate(bin_num = gsub("^14:[0-9][0-9]:00|^15:[0-9][0-9]:00", "14", x = bin_num))%>%    
  mutate(bin_num = gsub("^16:[0-9][0-9]:00|^17:[0-9][0-9]:00", "16", x = bin_num))%>%
  mutate(bin_num = gsub("^18:[0-9][0-9]:00|^19:[0-9][0-9]:00", "18", x = bin_num))%>%     
  mutate(bin_num = gsub("^20:[0-9][0-9]:00|^21:[0-9][0-9]:00", "20", x = bin_num))%>%
  mutate(bin_num = gsub("^22:[0-9][0-9]:00|^23:[0-9][0-9]:00", "22", x = bin_num))

str(clean_df)
clean_df$bin_num <- as.numeric(clean_df$bin_num)
```

# Regression Analysis

As discussed in the objective of this project, we will be using Logistic Regression classification algorithm to predict Hyperglycemia (high blood glucose) or Hypoglycemia (low blood glucose) in patients. 

These factors play an important role in blood glucose management and are directly associated with diet and exercise.

Through predicting the occurrence of these symptoms and the type of activities they are associated with, we can give recommendations to the patient.

******

**Model 1**

* In this section we will build a model that predicts hyperglycemic symptoms based on codes 

* The codes here represent different types of activities related to meal, exercise, insulin dose etc. (refer to the table from Description of Dataset)

* Before building the prediction model, we will spit the data into train and test dat frames with 70:30 ratio.

* Then we will create a confusion matrix and plot a ROC curve


Let us first encode the symptoms of BG

hyperglycemia = 1

Normal, Hypoglycemia = 0

```{r, message=FALSE, warning=FALSE, include=FALSE}

clean_df <- clean_df %>%
  mutate(sympcode = gsub("^([0-9]|[0-7][0-9]|80)$", 0, x = bg_conc))%>%
  mutate(sympcode = gsub("^(8[1-9]|9[0-9]|1[0-9][0-9])$", 0, x = sympcode))%>%
  mutate(sympcode = gsub("^(2[0-9][0-9]|3[0-9][0-9]|4[0-9][0-9]|50[0-1])$",
                         1, x = sympcode)) %>%
  mutate(sympcode = gsub("^(1.5|2.5|3.5|4.5|6.5|7.5)$", 0, x = sympcode))

str(clean_df)
clean_df$sympcode <- as.factor(clean_df$sympcode)
clean_df$bin_num <- as.numeric(clean_df$bin_num)
clean_df$activity <- as.factor(clean_df$activity)
str(clean_df)
```

First, let's find the accuracy of baseline model

```{r}
table(clean_df$sympcode)
```

24995 patients not hyperglycemic & 3851 are hyperglycemic

hence the accuracy of our baseline model is 87%

After building a binomial logistic regression model that Predicts Hyperglycemic symptoms  based on codes, a confusion matrix was created

```{r, message=FALSE, warning=FALSE, include=FALSE}
set.seed(844)
# split data into  70:30 ratio
split <- sample.split(clean_df$sympcode, SplitRatio = 0.7)
train <- subset(clean_df, split == TRUE)
test <- subset(clean_df, split == FALSE)
#create a model
```
```{r, echo=TRUE, message=FALSE, warning=FALSE}
symp.logit <- glm(sympcode~code+bin_num, data = train, family = "binomial")
summary(symp.logit) 
```
```{r, include=FALSE}
pred.test <- predict(symp.logit, type = "response", newdata = test)
summary(pred.test)
```

code 65 - 72 have negative coefficients and are not significant

Null deviance: 15872, Residual deviance: 11151, AIC: 11193

higher value in coefficients table are indicative of hyperglycemia

Confusion matrix, t = 0.1

```{r, echo=TRUE}
table(test$sympcode, pred.test > 0.1)
# accuracy of the model = (TP+TN)/(TP+TN+FP+FN)
(4555+1155)/(4555+1155+2944+1) # 0.66
```

Confusion matrix, t = 0.3

```{r, echo=TRUE}
table(test$sympcode, pred.test > 0.3)
```
accuracy = 0.79


Confusion matrix, t = 0.5

```{r, echo=TRUE}
table(test$sympcode, pred.test > 0.5)
```

accuracy = 0.86


Let's plot a **ROC curve** to pick a better threshold value 

```{r, echo=TRUE}
ROCRpred <- prediction(pred.test, test$sympcode)
ROCRperf <- performance(ROCRpred, "tpr", "fpr")
plot(ROCRperf, colorize = TRUE, print.cutoffs.at=seq(0,1,0.1), text.adj=c(-0.2,1.7))
```

Now to calulate **Area under curve AUC**

```{r, echo=TRUE}
as.numeric(performance(ROCRpred, "auc")@y.values)
```

Calculate the **F1 score/ harmonic mean**

```{r, echo=TRUE}
(2*7449)/(2*7449+1121+50)
```

**Results**

* Our model rarely predicts the risk of hyperglycemia above 50% and the accuracy of the model (86%) was very near to the accuracy of the baseline model (87%)

* Model can differentiate between patients who have probability of suffering from hyperglycemia (AUC = 83%)

* Note that the accuracy of our test is determined by F1 score. The value of which is 0.92 suggesting the strong accuracy of the model.

*******

**Model 2**

* In this section we will build a model that predicts blood glucose concentration based on activities (Meal)

* Activities are nothing but the similar codes grouped together into 5 bins
    + Insulin (code: 33 - 35)
    + Meal (code: 58 - 64, 66, 67, 68)
    + Exercise (code: 69 - 71 )
    + Unspecified (code: 48, 57, 72)
    + Hypoglycemic measurements (code: 65)
    
* Similar to model 1 we will split the data in 70:30 ratio into a train and test data

* Then we will create a confusion matrix and plot a ROC curve


Let's first add a new column and encode the activities

insulin, exercise, unspecified, hypoglycemic measurement = 0

meal = 1

```{r, message=FALSE, warning=FALSE, include=FALSE}
clean_df <- clean_df %>%
  mutate(act.code = gsub("^(3[3-5])$", 0, x = code)) %>%  # insulin = 0
  mutate(act.code = gsub("^(5[8-9]|6[0-4]|66|67|68)$", 1, x = act.code)) %>%   #meal = 1
  mutate(act.code = gsub("^(6[9]|7[0-1])$", 0, x = act.code)) %>% #exercise = 0
  mutate(act.code = gsub("^(48|57|72)$", 0, x = act.code)) %>% #unspecified = 0
  mutate(act.code = gsub("^(65)$", 0, x = act.code)) #hypo-measurement = 0

str(clean_df)
clean_df$act.code <- as.factor(clean_df$act.code)
```

```{r, echo=TRUE}
table(clean_df$act.code)
```

The accuracy of the baseline model is 61%

Now, let's create a binomial logistic regression model that Predicts BG concentration based on activities (meal)

```{r, message=FALSE, warning=FALSE, include=FALSE}
set.seed(844)
# split data into  70:30 ratio
act.split <- sample.split(clean_df$act.code, SplitRatio = 0.7)
act.train <- subset(clean_df, split == TRUE)
act.test <- subset(clean_df, split == FALSE)
```
```{r, echo=TRUE}
#create a model
act.logit <- glm(act.code~bg_conc, data = act.train, family = "binomial")
summary(act.logit) 
```
```{r, include=FALSE}
pred.test1 <- predict(act.logit, type = "response", newdata = act.test)
summary(pred.test1)
```

Confusion matrix, t = 0.3

```{r}
table(act.test$act.code, pred.test1 > 0.3)
```

accuracy = 0.65

Confusion matrix, t = 0.3

```{r}
table(act.test$act.code, pred.test1 > 0.5)
```

accuracy = 0.79

Let's plot a **ROC curve** to pick a better threshold value 

```{r, echo=TRUE}
ROCRact.pred <- prediction(pred.test1, act.test$act.code)
ROCRact.perf <- performance(ROCRact.pred, "tpr", "fpr")
plot(ROCRact.perf, colorize = TRUE, print.cutoffs.at=seq(0,1,0.1), text.adj=c(-0.2,1.7))
```

Now to calulate **Area under curve AUC**

```{r, echo=TRUE}
as.numeric(performance(ROCRact.pred, "auc")@y.values)
```

Calculate the **F1 score/ harmonic mean**

```{r, echo=TRUE}
(2*4679)/(2*4679+1136+640)
```

**Results**

* The model 2 has an accuracy of 79% which is better than our baseline model with accuracy 61%

* The AUC for this model is 88% and the value of F1 score is 0.84 which suggest the good accuracy of the model in prediction blood glucose concentration.

*******

**Model 3**

* In this section we will build a multinomial model that predicts Hypoglycemia or Hyperglycemia based on code

* We will split the data in 70:30 ratio into a train and test data

* After creating a confusion matrix for multinomial regression, we will calculat the classification and misclassification by the model

* Finally a 2-tailed z test will be performed to predict the probability of factors associating with blood glucode contration


Let's first add a new column and encode the symptoms

Normal = 1

Hypoglycemia = 2

Hyperglycemia = 3

Store these results in a new column "sympcode"

```{r, message=FALSE, warning=FALSE, include=FALSE}
clean_df <- clean_df %>%
  mutate(sympcode.mult = gsub("^([0-9]|[0-7][0-9]|80)$", 2, x = bg_conc))%>%
  mutate(sympcode.mult = gsub("^(8[1-9]|9[0-9]|1[0-9][0-9])$", 1, x = sympcode.mult))%>%
  mutate(sympcode.mult = gsub("^(2[0-9][0-9]|3[0-9][0-9]|4[0-9][0-9]|50[0-1])$",
                              3, x = sympcode.mult)) %>%
  mutate(sympcode.mult = gsub("^(1.5|2.5|3.5|4.5|6.5|7.5)$", 2, x = sympcode.mult))

clean_df$sympcode.mult <- as.factor(clean_df$sympcode.mult)
```

Now we will relevel this column; where Normal = 1

```{r, echo=FALSE, message=FALSE, warning=FALSE}
clean_df$out <- relevel(clean_df$sympcode.mult, ref = "1")
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
set.seed(844)
mult.split <- sample.split(clean_df$sympcode.mult, SplitRatio = 0.7)
mult.train <- subset(clean_df, split == TRUE)
mult.test <- subset(clean_df, split == FALSE)
```

Let's create a multinomial model to predict Normal BG, hypoglycemia & hyperglycemia based on code

```{r, echo=TRUE}
mult.mod <- multinom(out~code, data = mult.train)
summary(mult.mod)
```
```{r, include=FALSE}
p <- predict(mult.mod, newdata = mult.test)
```

Create Confusion matrix

```{r, echo=FALSE, message=FALSE, warning=FALSE}
table(p, mult.test$sympcode.mult)
```

**2-tailed z test**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
z <- summary(mult.mod)$coefficients/summary(mult.mod)$standard.errors
p <- (1 - pnorm(abs(z), 0 , 1)) * 2
p
```

**Results**

* The model 3 has the misclassification rate of only 22%. Which means that the model classifies the test data correctly for 78%

* The result of 2-tailed z test shows that the p value of codes 33, 34, 48, & 57 - 64 is almost zero. Hence the confidence level for these codes is very high. 

This suggests that the activities represented by these codes are responsible for developing symptoms of Hypoglycemia.
